The unphysicality of Hilbert spaces

### 2024-11-05 Articles/2023-HilbertQuantum/Submissions/Quantum Studies/HilbertQuantum.tex
HilbertQuantum: better discussion of rigged HS

To make this determination, we need to provide a standard to judge the physicality of a formal definition.\cite{redei_tension_2020, north_physics_2021} Suppose we are studying a physical system. A mathematical definition is \textbf{physical} if it properly characterizes the physical system in the given conditions. That is, if we are able to \textbf{justify on physical grounds} that the given mathematical definition is needed to capture and only capture a particular aspect of the physical system under the given condition. On the other hand, a mathematical definition is \textbf{unphysical} if it can be shown to require properties or operations that cannot have a physical counterpart. The question of this paper, then, is whether the use of Hilbert spaces to characterize quantum systems can be justified on physical grounds, from physical requirements. This is the task at hand.

Some mathematicians were disappointed that we didn't discuss in depth approaches like rigged Hilbert spaces or the algebraic approach to quantum mechanics. Do rigged Hilbert spaces spell out exactly what each element of the Gelfand triple describes and give a physical justification that tells us exactly which of those are the physically correct specifications? Do algebraic approaches give a physical justification as to why things we measure in a lab must be modeled by a $C^*$-algebra, and consequently allow us to determine whether the use of Hilbert spaces to model states can be given (or not given) a physical justification? To our knowledge, and to the knowledge of the mathematicians that approached us, they do not. This does not mean that these works are of no interest to a physical audience. It just means they are not going to help with the task at hand.

@@

Obviously, while the argument is ultimately physical, and it is therefore the job of a physicist to make these arguments, it has to be a physicist that is well-versed in foundational issues in mathematics. The justification, in fact, will need precise understanding of the mathematical definitions and their implication for both the physics and the math. Let us give an example. In a previous work we found that a set of physically distinguishable objects must be a $T_0$ second countable topological space where each open set represents a verifiable statement, a statement allowing a test that terminates in finite time if and only if the statement is true. This result tells us why functions in physics must be ``well-behaved'' (i.e.~topologically continuous in the right topology), why sets with cardinality greater than the continuum are unphysical (i.e.~they cannot be given a $T_0$ second countable topology) and why the Banach-Tarski paradox does not apply in physics (i.e. non-Borel sets are unphysical as they cannot be associated with an experimental test). This tells us that, at the very least, non-separable Hilbert spaces are unphysical as they have cardinality greater than the continuum.\cite{pittphilsci18363} This ties together ideas from point-set topology, logic, set theory, measure theory and computer science. Clearly, one needs to have at least a basic understanding of these fields to be able to follow the argument, and most degree programs in physics do not include typically discuss these topics in depth.\footnote{A previous reviewer conflated the completeness of the Hilbert space with the existence of a complete orthonormal basis of said space.} However, in the end, only a physicist can judge whether the original premise, the evidence-based requirement of physics, justifies the mathematical requirements.

To summarize, we are not interested in understanding whether Hilbert spaces have nice mathematical features for calculating or proving results. The fact that other tools have been developed to solve other mathematical problems is not relevant to this discussion. We are only interested in the question of mathematical representation of quantum states, \textbf{physical} entities we prepare in a lab. Something we produce, manipulate and, in the end, measure. What physical requirements are we capturing when we model them mathematically? Are Hilbert spaces capturing, and only capturing, physically tenable assumptions, or not?

>>>>

To make this determination, one needs a standard to judge the physicality of a formal definition\cite{redei_tension_2020, north_physics_2021} and, rather tellingly, physics tradition does not provide one. Here is the standard we follow in our work.\cite{aop-book}  Suppose we are studying a physical system. A mathematical definition is \textbf{physical} if it properly characterizes the physical system in the given conditions. That is, if we are able to \textbf{justify on physical grounds} that the given mathematical definition is needed to capture and only capture a particular aspect of the physical system under the given condition. On the other hand, a mathematical definition is \textbf{unphysical} if it can be shown to require properties or operations that cannot have a physical counterpart. The question of this paper, then, is whether the use of Hilbert spaces to characterize quantum systems can be justified on physical grounds, from physical requirements. This is the task at hand.

@@

If the problem of justifying the use of mathematical structures in physics is, ultimately, a physical problem, it also follows that it cannot be solved simply by a different mathematical approach. For example, according to the literature, (CITE) rigged Hilbert spaces were developed to give a mathematical foundation to Dirac’s bra-ket formalism, and be able to talk about eigenstates of position and momentum. The question is not whether these can be useful at a mathematical level to solve certain problem. They clearly are. The question is whether, on physical grounds, this is a good idea. Should we really treat continuous and discrete spectra the same, or realize that they are, on physical grounds, two rather different things? Should the fact that the energy spectrum is typically discrete for bound states and continuous for unbound states more seriously, and realize that there is something in the topology of the spectra that is rather physically important? Should we then make the different more prominent in our thinking, and not hide in mathematical details? Similarly, one can take the algebraic route and start with an algebra of observables. This, some argue, make some mathematical problems easier to solve. Again, no question about it. It still begs the question: why is it that, on physical grounds, observables should form a $C^*$-algebra? What, then, is the suitable physically meaningful state space? Moreover, note that in both those approaches Hilbert spaces still play a crucial role. If Hilbert spaces are physically problematic, these approaches are still relying on an object that is physically problematic.

While the argument is ultimately physical, and it is therefore the job of a physicist to make these arguments, it has to be a physicist that is well-versed in foundational issues in mathematics. The justification, in fact, will need precise understanding of the mathematical definitions and their implication for both the physics and the math. It is often the case that the mathematical detail informs the physics and shouldn't be swept under the rug.  Let us give an example. In a previous work we found that a set of physically distinguishable objects must be a $T_0$ second countable topological space where each open set represents a verifiable statement, a statement allowing a test that terminates in finite time if and only if the statement is true. This result tells us why functions in physics must be ``well-behaved'' (i.e.~topologically continuous in the right topology), why sets with cardinality greater than the continuum are unphysical (i.e.~they cannot be given a $T_0$ second countable topology) and why the Banach-Tarski paradox does not apply in physics (i.e. non-Borel sets are unphysical as they cannot be associated with an experimental test). This tells us that, at the very least, non-separable Hilbert spaces are unphysical as they have cardinality greater than the continuum.\cite{pittphilsci18363} This ties together ideas from point-set topology, logic, set theory, measure theory and computer science. Clearly, one needs to have at least a basic understanding of these fields to be able to follow the argument, and most degree programs in physics do not include typically discuss these topics in depth.\footnote{A previous reviewer conflated the completeness of the Hilbert space with the existence of a complete orthonormal basis of said space.} However, in the end, only a physicist can judge whether the original premise, the evidence-based requirement of physics, justifies the mathematical requirements.

To summarize, we are not interested in understanding whether Hilbert spaces have nice mathematical features for calculating or proving results. The fact that other tools have been developed to solve other mathematical problems is not relevant to this discussion. We are only interested in the question of mathematical representation of quantum states, which are physical entities we prepare in a lab. Something we produce, manipulate and, in the end, measure. What physical requirements are we capturing when we model them mathematically? Are Hilbert spaces capturing, and only capturing, physically tenable assumptions, or not?
